Late Pleistocene stone artefact technology at Mau A, Yen Bai, Norther Vietnam

### 2022-02-25 analysis/paper/paper.Rmd
Reordering of methods section, along with additional citations

  The outlines formed by the shape landmarks were then normalized for size, orientation, and rotation. We used Generalized Procrustes Analysis to strip the flakes in the sample dataset of size, orientation, and rotation, leaving only shape data. Generalized Procrustes Analysis (or GPA) is a statistical tool for taking shapes on a 2D plane and normalizing them so as to only examine the absolute shape of a set of landmarks. For an example of GPA being utilized to examine 2D landmarks in an archaeological context, refer to Jung & Woo (2019).

  Finally, the shape landmark values were analysed by Principal Components Analysis, a form of exploratory data analysis. Our method for carrying out the PCA on the shape landmarks was inspired by Theska et al. (2020), which also performed 2D landmark data analysis using PCA. Principal Component Analysis takes the initial set of data and transforms the initial set of variables into a smaller set of variables whist still preserving most of the variation seen in the initial data set. PCA biplots, which plot out the first two primary components, were created to visualize differences based upon secondary categorical variables, namely flake reduction categories, excavation unit, and mass cluster. The first two primary components formulated the individual points on biplot, while the secondary variables (flake reduction categories, excavation unit, and mass clusters) were represented by colored outlines on the biplots

@@

## References

@@

Jung, H., & Woo, E. J. (2019). Changes in mandibular ramus shape from the Neolithic to modern periods in Korea. International Journal of Osteoarchaeology, 29(4), 634–643. https://doi.org/10.1002/oa.2759

@@

>>>>

  The outlines formed by the shape landmarks were then normalized for size, orientation, and rotation. We used Generalized Procrustes Analysis to strip the flakes in the sample dataset of size, orientation, and rotation, leaving only shape data. Generalized Procrustes Analysis (or GPA) is a statistical tool for taking shapes on a 2D plane and normalizing them so as to only examine the absolute shape of a set of landmarks. An example of GPA being applied to stone tool landmarks is Okumura & Araujo (2014).

  Finally, the shape landmark values were analysed by Principal Components Analysis, a form of exploratory data analysis. Principal Component Analysis takes the initial set of data and transforms the initial set of variables into a smaller set of variables whist still preserving most of the variation seen in the initial data set. PCA has been used to investigate 2D (Riede & Pedersen, 2018; Hoggard et al., 2019; Radinović & Kajtez, 2021) and 3D (Archer et al., 2021) outlines of stone artefacts in the past. Our method for carrying out the PCA on the shape landmarks was inspired by Theska et al. (2020), which also performed 2D landmark data analysis using PCA. PCA biplots, which plot out the first two primary components, were created to visualize differences based upon secondary categorical variables, namely flake reduction categories, excavation unit, and mass cluster. The first two primary components formulated the individual points on biplot, while the secondary variables (flake reduction categories, excavation unit, and mass clusters) were represented by colored outlines on the biplots

@@

## References 
Archer, W., Djakovic, I., Brenet, M., Bourguignon, L., Presnyakova, D., Schlager, S., Soressi, M., & McPherron, S. P. (2021). Quantifying differences in hominin flaking technologies with 3D shape analysis. Journal of Human Evolution, 150, 102912. https://doi.org/10.1016/j.jhevol.2020.102912

@@

Hoggard, C. S., McNabb, J., & Cole, J. N. (2019). The Application of Elliptic Fourier Analysis in Understanding Biface Shape and Symmetry Through the British Acheulean. Journal of Paleolithic Archaeology, 2(2), 115–133. https://doi.org/10.1007/s41982-019-00024-6

@@

Okumura, M., & Araujo, A. G. M. (2014). Long-term cultural stability in hunter–gatherers: A case study using traditional and geometric morphometric analysis of lithic stemmed bifacial points from Southern Brazil. Journal of Archaeological Science, 45, 59–71. https://doi.org/10.1016/j.jas.2014.02.009

Radinović, M., & Kajtez, I. (2021). Outlining the knapping techniques: Assessment of the shape and regularity of prismatic blades using elliptic Fourier analysis. Journal of Archaeological Science: Reports, 38, 103079. https://doi.org/10.1016/j.jasrep.2021.103079

Riede, F., & Pedersen, J. B. (2018). Late Glacial Human Dispersals in Northern Europe and Disequilibrium Dynamics. Human Ecology, 46(5), 621–632. https://doi.org/10.1007/s10745-017-9964-8
